### 2018-02-01 submission/manuscript.Rmd
run spell-check

**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach. First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communites. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase. Finally, the chimera prevalence correlation with number of OTUs was assessed for both cycle number and HiFi DNA polymerase dependent differences.

@@

Over recent years there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, investigating ways that 16S rRNA gene surveys can be made more replicatable has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].  Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.

@@

***PCR Protocol:***
The five different high fidelity (HiFi) Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified. For Kappa and Q5, the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.

@@

***Sequence Processing:***
The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009]. Generally, the protocol followed what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data tranformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people. Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curits indicies were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). Based on these observations we wanted to next analyze potential reasons for these differences. First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each DNA polymerase. After assessing these errors, the total number of chimeras was determined after sequence processing. For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found in the SRA at the following accesssion number **need to upload still**.

@@

***Number of OTUs is Dependent on the HiFi DNA Polymerase Used:***
In order to compare the number of OTUs across individuals a Z-score normalization by individual, by cycle number, was employed on the number of OTUs data. After normalization, we identified that there was a HiFi DNA polymerase dependent difference that was consistent across sub-sampling [Figure 1]. Lower cycle numbers (15-20) tended to result in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 tended to have more distinct differences [Figure 1]. All subsampling levels were significantly different only at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1]. There were differences in HiFi DNA polymerase at 25 and 30 cycles but the sub-sampling depth had to be 5000 or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1]. Platinum HiFi DNA polymerase was the main driver of the differences observed across all sub-sampling depths at 35 cycles, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].

@@

***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
Overall, there was very little difference between each respective 5-cycle increment (e.g. 15x vs 20x) for both fecal and mock samples and this was consistent across the different sub-samplings used [Figure 3]. There were only two differences between 5-cycle increments that were identified. First, there were differences for the same fecal sample between 20x vs. 25x that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (subsampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), subsampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), subsampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A]. Second, for the mocks, where data is available, there were larger difference between 20x and 25x (subsampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B]. However, these differences between the next 5 cycle increment do not persist once 25 cylces are reached [Figure 3].

Using PERMANOVA to test for differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively). Phusion was one of only two HiFi DNA polymerases that managed to have samples for the 1000 sub-sampling depth at 15 cycles. Next, we assessed whether there were any major differences between 5 cycle increments within each sample. We found that there was no detectable difference in Bray-Curits index when comparing to the previous 5 cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05). However, Phusion at 1000 sub-sampling depth had a P-value = `r format(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000, taq == "PHU")[, "pvalue"], digits = 1, nsmall = 1)` before multiple comparison correction. It should be noted that at higher sub-sampling depths these differences in Bray-Curits indicies disappear for both differences in cycle number and within 5 cycle increments within an individual.

@@

The total seqeunces with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the pre.cluster step [Figure S1]. Major differences before this pre.cluster step were driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerase tested [Figure S1 & Table S7 & S8]. Although Accuprime^TM^ had the lowest per base error rate it had the largest number of sequences with at least one error, regardless of cycle number or sub-sampling depth [Figure S1]. However, this increased number of sequences with an error can be drastically lowered with existing bioinformatic approaches [Figure S1]. Investigation of whether there were HiFi DNA polymerase dependent effects on base subsitution found that there were generally no biases in the types of subsitution made [Figure S2].

@@

Although we did not observe strong differences in the Bray-Curtis index, based on cycle number, the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30x, that are commonly used. Additionally, there was no difference within individuals between corresponding 5 cycle increments (e.g. 15 to 20, 20 to 25, etc.). Conversely, we may not have enough samples to adequately judge the true magnitude difference between 20x and 25x communities and higher cycle numbers analyzed, even though there is a clear trend to suggest that 20x is very different then 25x [Figure 3]. This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cylces are reached. Increasing the sub-sampling depth, for some DNA polymerase, may reduce some of these observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5]. The chimera prevalence was strongly correlated with the number of OTUs which is relied upon heavily for many community metric calculations. However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase. It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully downweight these respective OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases (e.g. Platinum).

@@

The authors would like to thank all the study participants ERIN whose samples were utilized. We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan. Salary support for Marc Sze came from the Canadian Institue of Health Research and the Michigan Institute for Clinical and Health Research Postdoctoral Translational Scholar Program.

@@

**Figure 2:  Mock Sample Variability in Number of OTUs based on HiFi DNA Polymerase.** A) Sub-sampled to 1000 reads. B) Sub-sampled to 5000 reads. C) Sub-sampled to 10000 reads. The dotted line represents the number of OTUs genereated when the mock reference sequences are run through the pipeline.

>>>>

**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach. First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communities. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase. Finally, the chimera prevalence correlation with number of OTUs was assessed for both cycle number and HiFi DNA polymerase dependent differences.

@@

Recently there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced bias, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017].  Between the two approaches similar bias considerations include, but are not limited to, preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.

@@

***PCR Protocol:***
The five different high fidelity (HiFi) DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified. For Kappa and Q5, the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.

@@

***Sequence Processing:***
The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009]. Generally, the protocol followed what has been previously published [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithm instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people. Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curtis indices were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). Based on these observations we wanted to next analyze potential reasons for these differences. First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each DNA polymerase. After assessing these errors, the total number of chimeras was determined after sequence processing. For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017. The raw sequences can be found in the SRA at the following accession number **need to upload still**.

@@

***Number of OTUs is Dependent on the HiFi DNA Polymerase Used:***
In order to compare the number of OTUs across individuals a Z-score normalization by individual, by cycle number, was employed on the number of OTUs data. After normalization, we identified that there was a HiFi DNA polymerase dependent difference that was consistent across sub-sampling [Figure 1]. Lower cycle numbers (15-20) tended to result in less differences between HiFI DNA polymerase while cycle numbers of 25, 30, and 35 tended to have more distinct differences [Figure 1]. All sub-sampling levels were significantly different only at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001) [Table S1]. There were differences in HiFi DNA polymerase at 25 and 30 cycles but the sub-sampling depth had to be 5000 or higher (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "30x", sub_sample_level != 1000)[, "bh"]) < 0.05 & max(filter(fs_anova_summary, cycle == "25x", sub_sample_level != 1000)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S1]. Platinum HiFi DNA polymerase was the main driver of the differences observed across all sub-sampling depths at 35 cycles, based on a Tukey post-hoc test (P-value `r ifelse(max(filter(fs_tukey_summary, padj < 0.05)[, "padj"]) < 0.05,  invisible("<"), invisible("check data"))` 0.05) [Table S2].

@@

***Minimal Bray-Curtis Differences are Detected by Cycle Number:***
Overall, there was very little difference between each respective 5-cycle increment (e.g. 15x vs 20x) for both fecal and mock samples and this was consistent across the different sub-samplings used [Figure 3]. There were only two differences between 5-cycle increments that were identified. First, there were differences for the same fecal sample between 20x vs. 25x that was independent of HiFi DNA polymerase but dependent on sub-sampling depth (sub-sampled to 1000 = `r round(median(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`) (median (25% - 75% quantile)), sub-sampled to 5000 = `r round(median(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["5000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`), sub-sampled to 10000 = `r round(median(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(dist_data[["10000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3A]. Second, for the mocks, where data is available, there were larger difference between 20x and 25x (sub-sampled to 1000 = `r round(median(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"]), digits = 2)` (`r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["25%"], digits = 2)` - `r round(quantile(filter(mock_dist_data[["1000"]], cycle_compare == "20to25")[, "distance"])["75%"], digits = 2)`)) [Figure 3B]. However, these differences between the next 5 cycle increment do not persist once 25 cycles are reached [Figure 3].

Using PERMANOVA to test for differences within HiFi DNA polymerase groups based on cycle number, only Phusion had cycle dependent differences at 1000 and 5000 sub-sampling depth (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)` and `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 5000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`, respectively). Phusion was one of only two HiFi DNA polymerases that managed to have samples for the 1000 sub-sampling depth at 15 cycles. Next, we assessed whether there were any major differences between 5 cycle increments within each sample. We found that there was no detectable difference in Bray-Curtis index when comparing to the previous 5 cycle increment (P-value `r ifelse(min(bray_sim_to_prev_cycle_kruskal_results$bh, na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05). However, Phusion at 1000 sub-sampling depth had a P-value = `r format(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000, taq == "PHU")[, "pvalue"], digits = 1, nsmall = 1)` before multiple comparison correction. It should be noted that at higher sub-sampling depths these differences in Bray-Curtis indices disappear for both differences in cycle number and within 5 cycle increments within an individual.

@@

The total sequences with at least one error had multiple differences at different cycle numbers and was mostly alleviated by the use of the pre.cluster step [Figure S1]. Major differences before this pre.cluster step were driven by large differences in Accuprime^TM^ and Platinum versus the other HiFi DNA polymerase tested [Figure S1 & Table S7 & S8]. Although Accuprime^TM^ had the lowest per base error rate it had the largest number of sequences with at least one error, regardless of cycle number or sub-sampling depth [Figure S1]. However, this increased number of sequences with an error can be drastically lowered with existing bioinformatic approaches [Figure S1]. Investigation of whether there were HiFi DNA polymerase dependent effects on base substitution found that there were generally no biases in the types of substitution made [Figure S2].

@@

Although we did not observe strong differences in the Bray-Curtis index, based on cycle number, the data suggests that there may be differences between 15 and 20 cycles versus higher cycle numbers, such as 30x, that are commonly used. Additionally, there was no difference within individuals between corresponding 5 cycle increments (e.g. 15 to 20, 20 to 25, etc.). Conversely, we may not have enough samples to adequately judge the true magnitude difference between 20x and 25x communities and higher cycle numbers analyzed, even though there is a clear trend to suggest that 20x is very different then 25x [Figure 3]. This finding, in conjunction with the PERMANOVA results, suggest that cycle number can change bacterial community calculations but that these differences are minimal once 25 cycles are reached. Increasing the sub-sampling depth, for some DNA polymerase, may reduce some of these observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence differences between the different HiFi DNA polymerases [Figure 5]. The chimera prevalence was strongly correlated with the number of OTUs which is relied upon heavily for many community metric calculations. However, Bray-Curtis analysis with PERMANOVA showed few differences based on DNA polymerase. It is possible that many of the increased number of OTUs are not highly abundant allowing the Bray-Curtis index to be able to successfully down-weight these respective OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate these observed changes due to high chimera prevalence in specific HiFi DNA polymerases (e.g. Platinum).

@@

The authors would like to thank all the study participants ERIN whose samples were utilized. We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan. Salary support for Marc Sze came from the Canadian Institute of Health Research and the Michigan Institute for Clinical and Health Research Postdoctoral Translational Scholar Program.

@@

**Figure 2:  Mock Sample Variability in Number of OTUs based on HiFi DNA Polymerase.** A) Sub-sampled to 1000 reads. B) Sub-sampled to 5000 reads. C) Sub-sampled to 10000 reads. The dotted line represents the number of OTUs generated when the mock reference sequences are run through the pipeline.
